Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-04 paper/main.tex
started preparing popl submission

\documentclass[nocopyrightspace,preprint]{sigplanconf}

@@

@@

\title{Compositional Certified Resource Bounds}
% Certified Resource Bounds for C
% Compositional Certified Resource Bounds for C
% Compositional Quantitative Resource Analysis of C Programs
% Compositional Resource-Bound Certification for C Programs

\authorinfo
{Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}
{Yale University}

@@

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  The
  analysis framework consists of three parts.  First, the runtime
  resource consumption of programs is formalized with a parametric
  cost semantics for CompCert Clight.  Second, a quantitative Hoare
  logic that can be used for interactive bound development is
  implement and proved sound with respect to the cost semantics in the
  Coq Proof Assistant.  The logic is very expressive and shallow
  embedding enables a derived bound to be any function that is
  definable in Coq.  Third, an automatic amortized resource analysis
  for C programs that computes derivations in the quantitative Hoare
  logic is developed and implemented.  It is the first automatic
  amortized analysis that can derive bounds that depend on negative
  integers or differences between numbers, which is crucial to handle
  typical systems code.  Both, the quantitative logic and the
  automatic amortized analysis are naturally compositional and can be
  combined to semi-automatically derive global resource bounds.

  An experimental evaluation demonstrates the practicability of the
  analysis framework.  The expressivity of the logic is shown by
  manually deriving customized bounds that are tailored to specific
  algorithms.  A comparison of the automatic amortized analysis with
  other automatic tools on 30 challenging examples from the literature
  and open-source software shows that the bounds derived by the
  automatic amortized analysis are often more precise.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}

@@

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of the existing techniques in practice.
\begin{itemize}
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation--based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-3ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
%
In this work, we develop a resource analysis framework for C programs
that is based on a solid semantic foundation.  The choice of C is
primarily motivated by our ongoing work on the formally verified
hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
developed in C and is supposed to provide verified guaranties on
timing and memory usage.  Moreover, C is a natural choice because it
is still the most widely used language for system development; in
particular in safety-critical embedded and real-time systems where
resource bound analyses are often required by regulatory
authorities~\cite{May2013}.  Our \emph{contributions} are as follows.
%
\begin{enumerate}
\item We define a operational cost semantics for CompCert Clight that
  defines the resource consumption of terminating and diverging
  executions.  The cost is parametric in a user-definable cost metric
  and can be negative if resources are released during an execution.
\item We develop a quantitative Hoare logic for interactively deriving
  resource bounds for Clight programs.  The logic is implemented and
  proved sound in the Coq Proof Assistant with respect to the cost
  semantics of Clight.
\item We describe an automatic static analysis that computes bounds
  together with derivations in the quantitative Hoare logic.  The
  analysis is inspired by type-based amortized resource analysis for
  functional programs~\cite{Jost03,HoffmannAH11}.  It computes
  derivations by generating simple linear constraints that can be
  solved by an off-the-shelf LP solver and does not require any
  fixpoint computations to obtain loop invariants.
\item We show with a publically available prototype implementation and
  an experimental evaluation that our novel amortized resource
  analysis works precisely and efficiently for challenging loop and
  recursion patterns, and that the derived constant factors in the
  bounds are close to or identical with the optimal ones.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}
%
To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we directly derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can handle resources like memory
that may become available during the execution.

Our starting point is a recent work~\cite{veristack14} in which we
have formally verified bounds on the stack usage of C programs.  A key
part of this stack-bound verification is based on a quantitative Hoare
logic for deriving abstract stack bounds for CompCert Clight programs
that depend on the depth of (recursive) function calls.  % For
% non-recursive programs it is possible to compute derivations for such
% stack bounds in the quantitative Hoare logic automatically.  
% By compiling these bounds together with the C program, we have derived
% verified stack bounds for x86 assembly code.
%
Our \emph{first contribution} is a generalization of the quantitative
Hoare logic for CompCert Clight that enables us to derive resource
bounds that are parametric in the resource of interest.  % The
% quantitative logic is implemented in Coq and proved sound with respect
% to a small-step operational cost semantics for Clight.
% that counts a
% constant resource consumption for each atomic execution step.  The
% resource consumption at each step is parametric in a user-definable cost
% metric and can be negative to express the release of resources (as
% needed, e.g., to reason about memory usage).  Furthermore, a specific
% \emph{tick} function can be inserted in the source code to further
% customize the cost model.
In the implementation of the logic, we use
a shallow embedding in Coq which makes the logic very flexible.  There
is practically no limitation on the format of the derived resource
bounds since they can be any function that is definable in Coq.  The
derived bounds can also be parametric in a set of cost metrics or
specifically apply to one fixed cost metric.

@@

For functional programs, there exist resource analysis systems---based
on type-based amortized resource analysis~\cite{Jost03,
  HoffmannAH12}---that can automatically derive complex polynomial
bounds with tight constant factors.  This type-based amortized
resource analysis has been an inspiration in the design of the
quantitative Hoare logic and is therefore a natural candidate to
automate the reasoning.  However, it is a well-known problem that
type-based amortized resource analysis works well for functional
programs with pattern matching but not for C-like programs with
control flow that depends integer arithmetic.  For example, it is a
long-time open problem how to use type-based amortized analysis to
derive bounds for functions whose resource consumption depends on
(possibly negative) integers and non-sequential control flow as
introduced by \code{break} and \code{return} statements.
Unfortunately, these are the functions with non-constant resource
consumption that are most common in CertiKOS and other system
software.

Our \emph{main contribution} is an automatic amortized resource
analysis for CompCert Clight that computes derivations in the
quantitative Hoare logic.  It is the first automatic amortized
resource analysis that can derive bounds that depend on negative
integers and handles programs with mutually recursive functions as
well as \code{break} and \code{return} statements.  Using
potential-based approach of amortized analysis allows us to
automatically handle programs that are need special treatment in other
techniques such as
\begin{itemize}
\item interaction between \emph{sequential loops or function calls} through
  size changes of variables,
\item \emph{nested loops} that influence each others number of iterations,
\item and \emph{tricky iteration patterns} as found for instance in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
%
The main innovation that makes the analysis practical for C is the use
of interval sizes in potential functions instead of the sizes of
variables.  This leads to precise bounds of programs whose resource
consumption can be described as a function of sizes $|[x,y]| =
\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
  i=i+K)} where the step-size $K>0$ is a constant.
%

@@

\section{Overview and Examples}
\label{sec:inform}

@@

\section{Syntax and Semantics}
\label{sec:sem}

@@

\pref{lem:sem1} summarizes the main properties of the resource
counter of the semantics.  It can be proved by induction on the number
of steps.
\begin{lemma}
\label{lem:sem1}
  \begin{enumerate}
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
    \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
    \cont', d')$ then $\cost - \cost' = d - d'$.
  \end{enumerate}
\end{lemma}
%
If we assume that we have a \emph{positive resource metric}, that is,
a metric that assigns a positive cost to each evaluation step then we
can link bounds on the resource consumption under the metric to
termination.  This is formalized by \pref{lem:sem_term}.
%
\begin{lemma}
 \label{lem:sem_term}
 Let $M$ be a positive metric and assume that there exists $b \in
 \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
 (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
 Then there exists $B \in \N$ such that $(\state, S, \cont, d)
 \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
 and $\state''$ implies $n \le B$.
\end{lemma}

\section{Quantitative Hoare Logic}
\label{sec:logic}

@@

\section{Automatic Amortized Analysis}
\label{sec:AAA}

@@

Another interesting rule is {\sc Q:Call}.  It needs to account for the
changes to the stack caused by the function call, the argument/return
value passing, and the preservation of local variables.  We can sum up
the main ideas of the rule as follows.
\begin{itemize}
\item The potential in the pre- and postconditions of the function
  specification is equalized to its matching potential in the callee's
  pre/postcondition.
\item
  The potential of intervals $|\inter x y|$ is preserved
  if $x$ and $y$ are local variables,
\item
  The unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are set to zero in the postcondition.
\end{itemize}
If $x$ and $y$ are local variables and $\code{f}(x,y)$ is
called, the rule takes care to split the potential of
$|\inter x y|$ in two parts, one part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
realized by the equations $Q = P + S$ and $Q' =
P' + S'$.  Arguments in the function precondition
$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
of names different from all program variables.  This
prevents name conflicts from happening and ensures that
the substitution $[\Vargs/\vec x]$ is meaningful.
Symmetrically, we use the unique name $\Vret$ to represent
the return value in the function postcondition
$(\Gamma'_f, Q'_f)$.

@@

% \section{Implementation}

@@

% \section{Polynomial Potential}

@@

\section{Experimental Evaluation}
\label{sec:exper}

@@

\section{Related Work}

@@

\section{Conclusion}

@@

\bibliographystyle{abbrvnat}
\bibliography{lit}

>>>>

%Comment for anonymization
%\def\fullversion{}

\documentclass[nocopyrightspace]{sigplanconf}

\newcommand{\iffull}[2]{\ifx\fullversion\undefined{#2}\else{#1}\fi}
\newcommand{\ifshort}[2]{\ifx\fullversion\undefined{#1}\else{#2}\fi}

\newcommand{\itemskip}[0]{\ifshort{\vspace{-3pt}}{}}
\newcommand{\itemskipIn}[0]{\ifshort{\vspace{-1pt}}{}}
\newcommand{\sectskip}[0]{\ifshort{\vspace{-3pt}}{}}
\newcommand{\paraskip}[0]{\ifshort{\vspace{-2pt}}{}}
\newcommand{\lstskip}[0]{\ifshort{\vspace{-2pt}}{}}
\newcommand{\aftersectskip}[0]{\ifshort{\vspace{-1pt}}{}}
\newcommand{\subsectskip}[0]{\ifshort{\vspace{-1pt}}{}}

 % \usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
 %  \DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
 %  \captionsetup[figure]{format=myformat}
 % \renewcommand{\captionlabelfont}{\bf}

@@

% \abovedisplayskip=4pt
% \belowdisplayskip=4pt
% \abovedisplayshortskip=4pt
% \belowdisplayshortskip=4pt

@@

\title{Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}}
% Certified Resource Bounds for C
% Compositional Certified Resource Bounds for C
% Compositional Quantitative Resource Analysis of C Programs
% Compositional Resource-Bound Certification for C Programs

\authorinfo
{\iffull{Quentin Carbonneaux \and Jan Hoffmann \and Zhong Shao}{\vspace{-1cm}}}
{\iffull{Yale University}{\vspace{-1cm}}}

@@

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  The
  analysis framework consists of three parts.  First, the runtime
  resource consumption of programs is formalized with a parametric
  cost semantics for CompCert Clight.  Second, a quantitative Hoare
  logic that can be used for interactive bound development is
  implement and proved sound with respect to the cost semantics in the
  Coq Proof Assistant.  The logic is very expressive and shallow
  embedding enables a derived bound to be any function that is
  definable in Coq.  Third, an automatic amortized resource analysis
  for C programs that computes derivations in the quantitative Hoare
  logic is developed and implemented.  It is the first automatic
  amortized analysis that can derive bounds that depend on negative
  integers or differences between numbers, which is crucial to handle
  typical systems code.  Both, the quantitative logic and the
  automatic amortized analysis are naturally compositional and can be
  combined to semi-automatically derive global resource bounds.
  %
  An experimental evaluation demonstrates the practicability of the
  analysis framework.  The expressivity of the logic is shown by
  manually deriving customized bounds that are tailored to specific
  algorithms.  A comparison of the automatic amortized analysis with
  other automatic tools on 30 challenging examples from the literature
  and open-source software shows that the bounds derived by the
  automatic amortized analysis are often more precise.
\end{abstract}

@@

\sectskip
\section{Introduction}
\label{sec:intro}
\aftersectskip

@@

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of existing techniques in practice.
\itemskip
\begin{itemize}
\item Analysis tools are black boxes that \iffull{either}{} deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\itemskipIn
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\itemskipIn
\item Existing techniques often rely on complex external tools such as
  abstract interpretation--based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
\itemskip
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\itemskip
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-4ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
\itemskip
%
In this work, we develop a resource analysis framework for C programs
that is based on a solid semantic foundation.  The choice of C is
primarily motivated by \iffull{our}{the} ongoing work on the formally verified
hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
developed in C and is supposed to provide verified guaranties on
timing and memory usage.  Moreover, C is a natural choice because it
is still the most widely used language for system development; in
particular in safety-critical embedded and real-time systems where
resource bound analyses are often required by regulatory
authorities~\cite{May2013}.  Our \emph{contributions} are as follows.
%
\itemskip
\begin{enumerate}
\item We define a operational cost semantics for CompCert Clight that
  defines the resource consumption of terminating and diverging
  executions.  The cost is parametric in a user-definable cost metric
  and can be negative if resources are released during an execution.

\item We develop a quantitative Hoare logic for interactively deriving
  resource bounds for Clight programs.  The logic is implemented and
  proved sound in the Coq Proof Assistant with respect to the cost
  semantics of Clight.
\itemskipIn
\item We describe an automatic static analysis that computes bounds
  together with derivations in the quantitative Hoare logic.  The
  analysis is inspired by type-based amortized resource analysis for
  functional programs~\cite{Jost03,HoffmannAH11}.  It computes
  derivations by generating simple linear constraints that can be
  solved by an off-the-shelf LP solver and does not require any
  fixpoint computations to obtain loop invariants.
\itemskipIn
\item We show with a publically available prototype implementation and
  an experimental evaluation that our novel amortized resource
  analysis works precisely and efficiently for challenging loop and
  recursion patterns, and that the derived constant factors in the
  bounds are close to or identical with the optimal ones.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}
\itemskip
%
To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we directly derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can handle resources like memory
that may become available during the execution.

Our starting point is a recent work~\cite{veristack14} \iffull{in
  which we have}{that has} formally verified bounds on the stack usage
of C programs.  A key part of this stack-bound verification is based
on a quantitative Hoare logic for deriving abstract \emph{stack
  bounds} for CompCert Clight programs\ifshort{.}{ that depend on the
  depth of (recursive) function calls.}  % For
% non-recursive programs it is possible to compute derivations for such
% stack bounds in the quantitative Hoare logic automatically.  
% By compiling these bounds together with the C program, we have derived
% verified stack bounds for x86 assembly code.
%
Our \emph{first contribution} is a generalization of the quantitative
Hoare logic for CompCert Clight that enables us to derive resource
bounds that are parametric in the resource of interest.  % The
% quantitative logic is implemented in Coq and proved sound with respect
% to a small-step operational cost semantics for Clight.
% that counts a
% constant resource consumption for each atomic execution step.  The
% resource consumption at each step is parametric in a user-definable cost
% metric and can be negative to express the release of resources (as
% needed, e.g., to reason about memory usage).  Furthermore, a specific
% \emph{tick} function can be inserted in the source code to further
% customize the cost model.
In the implementation of the logic, we use
a shallow embedding in Coq which makes the logic very flexible.  There
is practically no limitation on the format of the derived resource
bounds since they can be any function that is definable in Coq.  The
derived bounds can also be parametric in a set of cost metrics or
specifically apply to one fixed cost metric.

@@

For functional programs, there exist resource analysis systems---based
on type-based amortized resource analysis~\cite{Jost03,
  HoffmannAH12}---that can automatically derive complex polynomial
bounds with tight constant factors.  This type-based amortized
resource analysis has been an inspiration in the design of the
quantitative Hoare logic and is therefore a natural candidate to
automate the reasoning.  However, it is a well-known problem that
type-based amortized resource analysis works well for functional
programs with pattern matching but not for C-like programs with
control flow that depends integer arithmetic.  For example, it is a
long-time open problem how to use type-based amortized analysis to
derive bounds for functions whose resource consumption depends on
(possibly negative) integers and non-sequential control flow as
introduced by \code{break} and \code{return} statements.
Unfortunately, these are the functions \iffull{with non-constant resource
consumption}{} that are most common in CertiKOS and other system
software.

Our \emph{main contribution} is an automatic amortized resource
analysis for CompCert Clight that computes derivations in the
quantitative Hoare logic.  It is the first automatic amortized
resource analysis that can derive bounds that depend on negative
integers. It also handles programs with mutually recursive functions as
well as \code{break} and \code{return} statements.  Using the
potential-based approach of amortized analysis allows us to
use a single mechanism to analyze programs that need special treatment in other
techniques.  For instance, we can naturally handle
\itemskip
\begin{itemize}
\item interaction between \emph{sequential loops or function calls} through
  size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each others number of iterations,
\itemskipIn
\item and \emph{tricky iteration patterns} as found for instance in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
\itemskip
%
The main innovation that makes the analysis practical for C is the use
of interval sizes in potential functions instead of the sizes of
variables.  This leads to precise bounds of programs whose resource
consumption can be described as a function of sizes $|[x,y]| =
\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
  i=i+K)} where the step-size $K>0$ is a constant.
%

@@

\sectskip
\section{Overview and Examples}
\label{sec:inform}
\aftersectskip

@@

\sectskip
\section{Syntax and Semantics}
\label{sec:sem}
\aftersectskip

@@

\pref{lem:sem1} summarizes the main properties of the resource
counter of the semantics.  It can be proved by induction on the number
of steps.
\begin{lemma}
\label{lem:sem1}
\itemskip
  \begin{enumerate}
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
    \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
\itemskipIn
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
    \cont', d')$ then $\cost - \cost' = d - d'$.
  \end{enumerate}
\itemskip
\end{lemma}
%
If we assume that we have a \emph{positive resource metric}, that is,
a metric that assigns a positive cost to each evaluation step then we
can link bounds on the resource consumption under the metric to
termination.  This is formalized by \pref{lem:sem_term}.
%
\begin{lemma}
 \label{lem:sem_term}
 Let $M$ be a positive metric and assume that there exists $b \in
 \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
 (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
 Then there exists $B \in \N$ such that $(\state, S, \cont, d)
 \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
 and $\state''$ implies $n \le B$.
\end{lemma}

\sectskip
\section{Quantitative Hoare Logic}
\label{sec:logic}
\aftersectskip

@@

\sectskip
\section{Automatic Amortized Analysis}
\label{sec:AAA}
\aftersectskip

@@

Another interesting rule is {\sc Q:Call}.  It needs to account for the
changes to the stack caused by the function call, the argument/return
value passing, and the preservation of local variables.  We can sum up
the main ideas of the rule as follows.
\itemskip
\begin{itemize}
\item The potential in the pre- and postconditions of the function
  specification is equalized to its matching potential in the callee's
  pre/postcondition.
\itemskipIn
\item
  The potential of intervals $|\inter x y|$ is preserved
  if $x$ and $y$ are local variables,
\itemskipIn
\item
  The unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are set to zero in the postcondition.
\end{itemize}
\itemskip
If $x$ and $y$ are local variables and $\code{f}(x,y)$ is
called, the rule takes care to split the potential of
$|\inter x y|$ in two parts, one part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
realized by the equations $Q = P + S$ and $Q' =
P' + S'$.  Arguments in the function precondition
$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
of names different from all program variables.  This
prevents name conflicts from happening and ensures that
the substitution $[\Vargs/\vec x]$ is meaningful.
Symmetrically, we use the unique name $\Vret$ to represent
the return value in the function postcondition
$(\Gamma'_f, Q'_f)$.

@@

% \sectskip
% \section{Implementation}

@@

% \sectskip
% \section{Polynomial Potential}

@@

\sectskip
\section{Experimental Evaluation}
\label{sec:exper}
\aftersectskip

@@

\sectskip
\section{Related Work}
\label{sec:related}
\aftersectskip

@@

\sectskip
\section{Conclusion}
\label{sec:concl}
\aftersectskip

@@

\sectskip
\bibliographystyle{abbrvnat}
\bibliography{lit}
